Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2021-03-18 manuscript.tex
added DNA-PAINT refs

Over the last of couple decades, DNA has proven to be much more than a vessel for genetic information. From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational works have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher-order dynamical processes and metastable intermediates. For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides.\citep{Sanstead2016} Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and how they relate to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

We find that 3spn2 enables access to long timescales in which we observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends. Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation. Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states. Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state behavior. Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process. Analysis within and between metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs. This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence. Taken together, our analysis reflects similar results to previous computational and experimental DNA studies, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths.\citep{Hinckley2013AnHybridization} Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients.\citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects} The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff and ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries. Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\citep{WuVariationalData, Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information. By comparing scores across several candidate feature sets (figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure. Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.

@@

The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein.\citep{Chen, Sidky}

@@

As an initial validation of 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments. Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (figure \ref{fig:relaxation-comparison})). At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

% A documented short-coming for the 3spn2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}. This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.

The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.

For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process. In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater enthalpic barrier associated with G:C fraying. We observe similar trends in the experimental data, although comparisons at high temperatures were limited due to mixing between the fast and slow responses (figure \ref{fig:relaxation-comparison}). We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3spn2 fraying dynamics are accelerated by about 100-200x relative to experiments. It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom. In particular, the simplified treatment of the fraying process may smoothen the free energy landscape and speed up dynamics relative to a more global processes like dissociation. Although one should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al. and that 3spn2 can capture some sequence-dependent fraying effects.
% need a better transition here back into SRV-MSMs?

@@

% potentially get rid of State representation, ask Brennan if they are meaningful?
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Fig4.pdf}
        \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate. Arrows indicate the probability of transitioning out of the state within the lag time. Circle areas are proportional to state stability.}

@@

% rewrite this into one paragraph
\subsection{MSM stabilities agree with nearest neighbor thermodynamic models}

>>>>

Over the last couple of decades, DNA has proven to be much more than a vessel for genetic information. From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} Recently, single molecule localization techniques have exploited the rapid and transient binding of short DNA oligomers in order to achieve super-resolution microscopy and optical multiplexing. \cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Indeed, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational works have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher-order dynamical processes and metastable intermediates. For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally and appears highly sequence-dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides.\citep{Sanstead2016} Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and how they relate to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3SPN2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

We find that 3SPN2 enables access to long timescales in which we observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends. Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation. Consistent with previous studies, \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states. Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state behavior. Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process. Analysis within and between metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs. This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence. Taken together, our analysis reflects similar results to previous computational and experimental DNA studies, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales. These can results inform a generalized model between DNA sequence and hybridization mechanism with application towards equilibrium binding technologies such as DNA-PAINT and barcoding.

@@

The 3SPN2 coarse-grained model was designed to accelerate DNA simulations relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths.\citep{Hinckley2013AnHybridization} Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients.\citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects} The 3SPN2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble with a Langevin thermostat that modeled interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries. Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinate construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\citep{WuVariationalData, Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information. By comparing scores across several candidate feature sets (Figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure. Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.

@@

The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features per frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein.\citep{Chen, Sidky}

@@

As an initial validation of 3SPN2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments. Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3SPN2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4$^{\circ}$C between experiment and simulation fits. While 3SPN2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (Figure \ref{fig:relaxation-comparison})). At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

% A documented short-coming for the 3SPN2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}. This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.

The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3SPN2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.

For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process. In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater enthalpic barrier associated with G:C fraying. We observe similar trends in the experimental data, although comparisons at the highest temperatures were limited due to mixing between the fast and slow responses (Figure \ref{fig:relaxation-comparison}). We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3SPN2 fraying dynamics are accelerated by about 100-200x relative to experiment. It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom. In particular, the simplified treatment of the fraying process may smoothen the free energy landscape and speed up dynamics relative to a more global processes like dissociation. Although one should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al. and that 3SPN2 can recapitulate some sequence-dependent fraying effects.
% need a better transition here back into SRV-MSMs?

@@

% potentially get rid of State representation, ask Brennan if they are meaningful?
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Fig4.pdf}
        \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate. Arrows indicate the probability of transitioning between states within the lag time. Circle areas are proportional to the log of stationary probabilities to enhance visualization of less stable states}

@@

% rewrite this into one paragraph
\subsection{Inert tails produce deviations from nearest neighbor thermodynamic models}
